---
title: "development of housing pressure and residential vacancy in Dutch municipalities between 2020 and 2025, and understanding coexistence and scarcity through price pressure, rental structure and urban density."
author: "[Marline Human (2860665)], [Storm Bakker (2852941)], [Nando Fredricks (2888962)], [Student (studentnummer)] and [Jermaine Koreman (2866400)]."
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Include your names: Marline Human, Nando Fredricks, Storm bakker, Jasper Haavekost en Jermaine Koreman.

Include the tutorial group number group: 1.5

Include your tutorial lecturer's name: Include your tutorial lecturer's name: Chantal Schouwenaar \>\>\>\>\>\>\> 85d130eb228492501734ba45d46a227b4a6605b6

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

The Dutch housing market is under unprecedented pressure. Across the Netherlands, rental prices have risen sharply, buying a home has become unaffordable for many households, and waiting lists for social housing continue to grow (CBS, 2025; Kences, 2025). At the same time, a paradox emerges: despite the apparent shortage, a significant number of dwellings remain unoccupied in many municipalities.

This combination of housing pressure and vacancy creates a social problem that does not affect all of the Netherlands equally. In some municipalities, housing pressure is severe and vacancy is low; in others, vacancy rates are surprisingly high despite national shortages. This spatial variation means that the opportunity to find affordable housing increasingly depends on where one lives — a form of geographic inequality that affects young adults, lower-income households, and people seeking to relocate.

### Why is this relevant?

This problem is relevant for several reasons:

**1. Economic relevance.** The housing market is a cornerstone of the economy. Rising housing costs reduce disposable income, limit labor mobility, and contribute to wealth inequality between homeowners and renters (CBS, 2025). Geographic differences in housing pressure also shape patterns of regional economic development.

**2. Social relevance.** Access to affordable housing is widely considered a basic right. When access depends heavily on geography, spatial inequality deepens and social cohesion is undermined. The coexistence of shortage and vacancy raises difficult questions about the efficiency and fairness of the housing system.

**3. Policy relevance.** Understanding where housing pressure is most severe — and where it coincides with vacancy — helps policymakers target interventions more effectively. A municipality-level analysis can inform decisions about new construction, rent regulation, and vacancy policies, and shows which areas need attention most urgently.

**4. Personal relevance.** As students at VU Amsterdam, we experience the consequences of housing pressure directly. By analyzing the problem at a national scale, we move beyond our own situation and investigate the structural forces shaping housing access across the Netherlands.

By quantifying and visualizing how housing pressure and vacancy vary across Dutch municipalities between 2020 and 2025, this project aims to make the scale and distribution of the problem visible, and to identify which types of municipalities face the sharpest paradox between scarcity and unused capacity.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
dataset <- midwest
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
